Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-09-25 paper/paper_draft.qmd
Proof reading of paper for checks and reducing word count
- restructured to move DHW calculation methods and qualitative bioregion scale clustering results to supplementary material
- fixed figure captions that have repetition

Climate change is impacting species survival, persistence and ecosystem functions globally.
This is particularly true for coral reefs, which are vulnerable to increasing magnitude and
variation in environmental stressors such as ocean temperature [@foster2021; @gattuso2014].
Mass bleaching and associated mortality events driven by climate change can cause
reef-building coral population declines [@hughes2017; @hughes2018; @bozec2025], possibly
leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011].
Climate impacts on coral reefs are decreasing the ability of coral reefs to provide
ecological functions such as benthic habitat and nutrient cycling [@forster2021;
@williams2019] as well as ecosystem services such as food, protection, tourism and
Indigenous cultural value [@eddy2021; @stoeckl2021].

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes. Coral
reef ecosystem resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - is conceptualised to depend on three interconnected mechanisms. First, the
ability of individual corals to survive thermal stress. Second, the connectivity networks
that enable recolonisation after disturbances [@hock2017; @mumby2008; @torres2024]. Third,
the maintenance, and accretion, of the carbonate framework that provides reef structure
[@perry2012]. We aim to assess how these mechanisms interact under projected climate
conditions to identify the characteristics of resilient reefs through scenario-based coral
population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@cappo2000].
The GBR faces severe projected coral cover declines under climate change [@bozec2025;
@bozec2022; @condie2021]. These coral cover declines will be compounded by ocean
acidification, tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012;
@fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass
bleaching event may be signalling a shift towards this reef future, with some level of
bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs)
[@australianinstituteofmarinescience2024; @cantin2024].

@@

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018]. For GBR mid-shelf and offshore reefs, maintaining
positive carbonate budgets requires approximately 17% live coral cover [@desbiens2021].

Recent advances in whole-ecosystem larval connectivity modelling **\[Insert GBR1
connectivity citation.\]** and coral thermal response research [@bairos-novak2021] enable
integrated investigation of how multiple factors interact to determine climate resilience.
By incorporating the latest coral population models [@ribeirodealmeida2024], ecosystem-wide
connectivity data, and updated climate projections [@dixon2022], we can examine which
factors most strongly influence reef persistence under different scenarios. While
connectivity has been emphasized as critical for GBR resilience [@hock2017], and remains an
important consideration, our integrated modelling reveals depth may play a fundamental role
in determining maintenance of reef coral populations and positive carbonate budgets under
current projections of climate change.

@@

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority (GBRMPA)
[@GBRMPA2022]. The reef definitions define 3806 reefs across the GBR. An indicative map of
the GBR marine park including the Marine Park Management areas [@GBRMPA2007] and the
corresponding bioregion for each reef is shown in (@fig-region-map; [@GBRMPA2024]).
Estimates of coral carrying capacity for each reef were taken from prior analyses undertaken
in [@bozec2022].

@@

These GCMs were selected based on evaluation conducted by Grose et al., [-@grose2023],
indicating their ability to recreate historical atmospheric temperature and precipitation
conditions in the Australasian region. They span a range of Equilibrium Climate Sensitivity
(ECS) values within the Intergovernmental Panel on Climate Change (IPCC) assessed likely
range. The IPCC defines ECS as the long-term warming caused by a doubling of carbon dioxide
above its pre-industrial concentration [@forster2021]. While this assessment does not
consider oceanic data, Grose et al. [-@grose2023] provide a region-specific selection
mechanism which we use as an indicator of general performance of GCMs for the region. An
overview of the selected GCMs is provided in @tbl-gcms. Note here that Grose et al.,
[-@grose2023] assessed GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023] provides
data based on GFDL-CM4. This substitution is assumed appropriate given they share the same
fundamental representation of the physical climate system.

Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the plausible emission
pathway. It represents a moderate emissions scenario with continued CO~2~ emissions
comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.66°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of these GCMs
therefore allow for a range of possible climate futures to be explored, capturing
uncertainty in future responses to CO~2~ increases (@sfig-ecs; @fig-methods-dhws;
[@grose2023; @shepherd2018]) and enables the assessment of modelled coral reef behaviour
over a broad range of independently derived climate futures [@abramowitz2019].

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of HotSpot SST Anomaly values that are greater than 1°C, expressed in units of
°C-weeks. The 12-week accumulation period reflects the cumulative nature of coral thermal
stress. Brief temperature spikes may be tolerable, but sustained or repeated exposure to
elevated temperatures causes progressively greater physiological stress and bleaching
severity. DHW values integrate both the magnitude and duration of thermal stress: a moderate
anomaly (e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly
(e.g., 3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for
each indicative reef (see @GBRMPA2022) were then determined by taking the mean DHW value for
all pixels touching each reef outline. The process described here was conducted for all 2197
reefs to create DHW projections spanning from 2025 to 2075. Code used to calculate DHW data
and perform analyses involved in this investigation can be found at Grier and Iwanaga
[-@grier2025].

@@

Rather than attempting to forecast precise outcomes, the exploratory approach aims to
systematically search many possible futures to identify patterns that lead to outcomes of
interest. The approach allows for the identification of robust management strategies that
perform well across a diversity of future conditions rather than attempting to identify a
single most optimal strategy. Concurrently, assessing the wide swathe of scenarios allows
contextual understanding of the conditions in which strategies may fail. Multiple
interacting and compounding sources of uncertainty, and a lack of broad agreement as to
their likelihood, create inherently uncertain outcomes. Here, exploratory modelling was
applied to identify the environmental and reef characteristics, such as thermal stress,
depth, and connectivity, that support favourable outcomes in reef recovery and resilience.
Metrics used in analyses can be found in @stbl-metrics.

All data preparation, modelling and analyses were performed using the Julia programming
language (The Julia Project; https://julialang.org/; [@bezanson2017]), with visualisations
created using the Makie plotting ecosystem [@danisch2021].

@@

Larval transition probability matrices were generated using the EReefs Great Barrier Reef
1km resolution hydrodynamic model and particle tracking simulations. Connectivity data from
24 days in total were used in this investigation to estimate coral larval flows. These data
span six years from 2015 to 2020. The years 2015, 2016, 2018 and 2019 are each represented
by connectivity matrices covering spawning periods over three days in each year (12 days in
total), while 2017 and 2020 are each represented by six spawning days each, coming to a
total of 24 days of connectivity matrices. The values in these matrices indicate the
probability of larval transfer between each pair of reefs in the GBR, herein referred to as
connectivity strengths. Indicative connectivity strengths are implemented in ADRIA-CoralBlox
as the mean of all available data for each location (i.e., mean of all 24 spawning days).
While the connectivity strengths do not vary from year to year within each ADRIA-CoralBlox
simulation, they do vary between reefs which are of varying sizes, providing sufficient
heterogeneity to capture spatial patterns in larval dispersal and settlement dynamics. This
spatial variability also enables investigations into the influence of connectivity strength
and its role in driving recovery processes.

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef (excluding retained larvae from the target
reef) and the mean timeseries absolute coral cover (km^2^) of each matching source reef.
This metric provides an approximation of overall larval flow into a reef, with high values
indicating the target reef is supplied high volumes of larvae via strong connections and low
values indicating the target reef has weak (or no) incoming connections or is supplied by
small source populations. A limitation of weighted incoming connectivity is it does not
incorporate the proportion of the source reef's population that is reproducing and producing
larvae, instead taking the total coral cover area as an indicator of larval productivity.
Outgoing connection strength is calculated as the sum of all outgoing transition
probabilities from a target reef. Outgoing connection strength has a theoretical maximum of
one (all larvae reach settlement on sink reefs), however the large size of the GBR combined
with larval mortality considerations result in outgoing connection strength values much
lower than one. The ratio of source-to-sink connectivity ($O:I$) for a target reef
represents the strength of the reef's outgoing larval flow relative to it's incoming larval
flow. This is calculated as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where S represents the sum of
connectivity probabilities, N represents the number of connections, O denotes outgoing
connectivity from the target reef and I denotes incoming connectivity to the target reef. If
a reef has no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming
connections in this calculation, do not directly represent the larval supply gained as the
breeding populations from which connections are coming from are not represented. A
source-to-sink ratio greater than 1 (a value greater than 0 on a log scale) indicates a reef
has greater outgoing than incoming larval flow. Reef larval retention probability indicates
the proportion of spawned larvae from a target reef that remain to settlement.

@@

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median time series for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour.

@@

The time series clustering was applied over the time frame between 2025 and 2075. The
clustering was conducted for three spatial scales: (1) the whole of GBR; (2) each GBR
management area; and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). The
analyses was conducted at the different spatial scales as, conceptually, the temporal
behaviour constituting "low", "medium" or "high" coral cover may differ at the different
spatial scales. As it was, analyses at the whole-of-GBR scale were found to lead to
identical conclusions regarding reef characteristics which influence coral cover and so for
brevity are not shown here.

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, outgoing connectivity, and log10-transformed
outgoing-to-incoming connectivity ratio. The full dataset consisted of the time series
cluster assignment of each of the 2197 reefs under the five GCMs explored, for a total of
10,985 samples. The dataset was randomly split into training (60%) and testing (40%) subsets
to train and assess classification performance. The RF classifier achieved 80% accuracy on
the training set and 60% on the test set.

A 3-fold cross-validation was also conducted, arriving at a similar 59% accuracy rate.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.70, 0.43 and 0.64 respectively, with precision of 0.62, 0.51 and 0.66 when evaluated with
the test set (consisting of 4394 samples) [@stbl-confusion-matrix;
@stbl-rf-cluster-performance]. Lower performance with the Medium cluster reflects the
inherent difficulty of classifying transitional states in natural systems. The consistent
level of accuracy across test and cross-validation demonstrates the generalised performance
of the RF classifier, while not accurate, was deemed sufficiently informative for the
purpose of assessing the influence of reef characteristics.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically varied (one-at-a-time) across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. Use of the median predicted
probability was also assessed but was found not to substantially influence the
interpretation of results (and so are not shown here). The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and incoming connectivity, two-way partial dependence
analysis was performed on grids along the range of these two features.

@@

In contrast to the broad temporal behaviour at the Management area scale, the relationship
between reef characteristics and assigned clusters were found to be similar at the bioregion
scale across GCMs. Results are only shown and discussed for ACCESS-ESM1-5, an intermediate
ECS scenario (@sfig-ecs). Those interested in results for the other assessed GCMs are
referred to Grier and Iwanaga [-@grier2025]. Across bioregions, reefs in the High cluster
not only had greater levels of coral recovery but displayed persistence to 2050
(@sfig-access-esm1-5-bioregion-timeseries). Reefs in the High cluster were consistently
found to have the greatest median reef depths across bioregions
(@fig-access-esm1-5-bioregion-depth).

![Median depth for reefs within each bioregion under ACCESS-ESM1-5, separated into clusters
assigned by time series clustering. Time series were clustered based on their behaviour
between 2030 and 2060. Time series with comparatively low median area-relative cover is
indicated in green, intermediate cover in orange, and high in blue. Solid lines represent
the median depth for each cluster across reefs and bands represent the distribution of depth
values within each
cluster.](../figs/ACCESS-ESM1-5/bioregion/depth_bioregion_violin.png){#fig-access-esm1-5-bioregion-depth
fig-align="center" width="14.82cm"}

Across bioregions, there were no consistent patterns between reef clusters in mean DHW
values (@sfig-access-esm1-5-bioregion-dhwmean), log10-weighted incoming connectivity
(@sfig-access-esm1-5-bioregion-weightedincoming), log10 outgoing connection strength
(@sfig-access-esm1-5-bioregion-outgoingconn), log10 source-to-sink connectivity ratio
(@sfig-access-esm1-5-bioregion-logsosi) and log10 larval retention probability
(@sfig-access-esm1-5-bioregion-selfconn). Across six of the 29 bioregions reefs in the high
cover cluster have lower log10 carrying capacity values than reefs in other clusters
(@sfig-access-esm1-5-bioregion-carrying-capacity).

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its assignment into
Low, Medium, or High cover clusters. Reef median depth, carrying capacity and mean time
series DHW level have the highest feature importance scores, accounting for 29%, 15% and 15%
predictive importance respectively (@fig-ts-cluster-pdp; Panel A). Area weighted incoming
connectivity, outgoing connectivity strength, source-to-sink connectivity ratio and larval
retention probability account for \~12%, 11%, 10% and 8% respectively.

@@

![Reef Degree Heating Week time series clustered within each management area, for each GCM
analysed. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover
between 2030 and 2060 within each bioregion. ‘Medium’ clusters (orange) are clusters with
intermediate median area-relative cover between 2030 and 2060 within each bioregion. ‘High’
clusters (blue) are clusters with the highest median area-relative cover between 2030 and
2060 within each bioregion. Solid lines represent the cluster median time series and bands
represent the 2.5 to 97.5 percentile confidence intervals. Each management area is
represented five times (for each GCM), with each GCM displayed
horizontally.](../figs/GCM_dhw_timeseries_plot.png){#sfig-gcm-dhw-trajectories
fig-align="center" width="14.82cm"}

@@

@@

![Reef area-relative coral cover time series clustered within each bioregion. ‘Low’ clusters
(green) are clusters with the lowest median area-relative cover between 2030 and 2060 within
each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median
area-relative cover between 2030 and 2060 within each bioregion. ‘High’ clusters (blue) are
clusters with the highest median area-relative cover between 2030 and 2060 within each
bioregion. Solid lines represent the cluster median time series and bands represent the 95th
percentile confidence intervals. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/bioregion_cover_timeseries.png){#sfig-access-esm1-5-bioregion-timeseries
fig-align="center" width="14.82cm"}

![Mean time series DHW values for reefs within each bioregion, separated into clusters
assigned by time series clustering. ‘Low’ clusters (green) are clusters with the lowest
median area-relative cover between 2030 and 2060 within each bioregion. ‘Medium’ clusters
(orange) are clusters with intermediate median area-relative cover between 2030 and 2060
within each bioregion. ‘High’ clusters (blue) are clusters with the highest median
area-relative cover between 2030 and 2060 within each bioregion. Solid lines represent the
median DHW level for each cluster and bands represent the distribution of mean DHW levels
within each cluster. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean
fig-align="center" width="14.82cm"}

>>>>

Climate change is impacting species survival, persistence and ecosystem functions on coral
reefs, which are vulnerable to increasing magnitude and variation in environmental stressors
such as ocean temperature [@foster2021; @gattuso2014]. Mass bleaching and associated
mortality events driven by climate change can cause reef-building coral population declines
[@hughes2017; @hughes2018; @bozec2025], possibly leading to coral-macroalgae regime shifts
[@arif2022; @wild2011]. Climate impacts on coral reefs are decreasing the ability of coral
reefs to provide ecological functions such as benthic habitat and nutrient cycling
[@forster2021; @williams2019] and services such as food, protection, tourism and Indigenous
cultural value [@eddy2021; @stoeckl2021].

Coral reef ecosystem resilience - the capacity to maintain coral populations and avoid
ecosystem regime shifts - is conceptualised to depend on three interconnected mechanisms.
First, the ability of individual corals to survive thermal stress. Second, the connectivity
networks that enable recolonisation after disturbances [@hock2017; @mumby2008; @torres2024].
Third, the maintenance, and accretion, of the carbonate framework that provides reef
structure [@perry2012]. We aim to assess how these mechanisms interact under projected
climate conditions to identify the characteristics of resilient reefs through scenario-based
coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs interconnected through larval dispersal patterns mediated by
hydrodynamic, spatial, and ecological processes [@cappo2000]. The GBR faces severe projected
coral cover declines under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover
declines will be compounded by ocean acidification, tropical cyclones, and crown-of-thorns
starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011;
@wolff2018]. The 2023-24 Summer mass bleaching event may be signalling a shift towards this
reef future, with some level of bleaching recorded on 79% of surveyed reefs (total 1001
surveyed reefs) [@australianinstituteofmarinescience2024; @cantin2024].

@@

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018].

Recent advances in whole-ecosystem larval connectivity modelling **\[Insert GBR1
connectivity citation.\]** and coral thermal response research [@bairos-novak2021] enable
integrated investigation of how multiple factors interact to determine climate resilience.
By incorporating the latest coral population models [@ribeirodealmeida2024], ecosystem-wide
connectivity data, and updated climate projections [@dixon2022], we can examine which
factors most strongly influence reef persistence under different scenarios. While
connectivity has been emphasized as critical for GBR resilience [@hock2017], and remains an
important consideration, our integrated modelling reveals depth plays a fundamental role in
determining maintenance of reef coral populations and positive carbonate budgets under
current projections of climate change.

@@

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA) [@GBRMPA2022]. The reef
definitions define 3806 reefs across the GBR. An indicative map of the GBR marine park
including the Marine Park Management areas [@GBRMPA2007] and the corresponding bioregion for
each reef is shown in [@fig-region-map] [@GBRMPA2024]. Estimates of coral carrying capacity
for each reef were taken from prior analyses undertaken in Bozec et al. [-@bozec2022].

@@

These GCMs were selected based on evaluation by Grose et al., [-@grose2023], indicating
their ability to recreate historical atmospheric temperature and precipitation conditions in
the Australasian region. They span a range of Equilibrium Climate Sensitivity (ECS) values
within the Intergovernmental Panel on Climate Change (IPCC) assessed likely range. The IPCC
defines ECS as the long-term warming caused by a doubling of carbon dioxide above its
pre-industrial concentration [@forster2021]. While this assessment does not consider oceanic
data, Grose et al. [-@grose2023] provide a region-specific selection mechanism which we use
as an indicator of general performance of GCMs for the region. An overview of the selected
GCMs is provided in @tbl-gcms. Note here that Grose et al., [-@grose2023] assessed
GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023] provides data based on GFDL-CM4.
This substitution is assumed appropriate given they share the same fundamental
representation of the physical climate system.

Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the plausible emission
pathway. It represents a moderate emissions scenario with continued CO~2~ emissions
comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.66°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of these GCMs
therefore allow for a range of possible climate futures to be explored, capturing
uncertainty in future responses to CO~2~ increases (@sfig-ecs; @fig-methods-dhws;
[@grose2023; @shepherd2018]) and enables the assessment of modelled coral reef behaviour
over a broad range of independently derived climate futures [@abramowitz2019; @brunner2020].

Daily SST data [@dixon2023] were used to calculate climatological baselines and “HotSpot
Anomaly” values for each pixel following NOAA methodology [@liu2017; @skirving2020].
Additional information about DHW calculation methods is provided in the supplemental
material. DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the
12-week rolling sum of HotSpot SST Anomaly values that are greater than 1°C, expressed in
units of °C-weeks. The 12-week accumulation period reflects the cumulative nature of coral
thermal stress. Brief temperature spikes may be tolerable, but sustained or repeated
exposure to elevated temperatures causes progressively greater physiological stress and
bleaching severity. DHW values integrate both the magnitude and duration of thermal stress:
a moderate anomaly (e.g., 1°C) sustained for 12 weeks produces the same DHW value as a
severe anomaly (e.g., 3°C) lasting 4 weeks, though the biological impacts may differ. DHW
estimates for each indicative reef (see @GBRMPA2022) were then determined by taking the mean
DHW value for all pixels touching each reef outline. The process described here was
conducted for all 2197 reefs to create DHW projections spanning from 2025 to 2075.

@@

Rather than attempting to forecast precise outcomes, the exploratory approach aims to
systematically search many possible futures to identify patterns that lead to outcomes of
interest. The approach allows for the identification of robust management strategies that
perform well across a diversity of future conditions rather than attempting to identify a
single most optimal strategy. Concurrently, assessing the wide swathe of scenarios allows
contextual understanding of the conditions in which strategies may fail. Multiple
interacting and compounding sources of uncertainty, and a lack of broad agreement as to
their likelihood, create inherently uncertain outcomes. Here, exploratory modelling was
applied to identify the environmental and reef characteristics, such as thermal stress,
depth, and connectivity, that support favourable outcomes in reef recovery and resilience.

All data preparation, modelling and analyses were performed using the Julia programming
language (The Julia Project; https://julialang.org/; [@bezanson2017]), with visualisations
created using the Makie plotting ecosystem [@danisch2021]. Code used to process DHW data and
perform analyses involved in this investigation can be found at Grier and Iwanaga
[-@grier2025].

@@

Larval transition probability matrices were generated using the EReefs Great Barrier Reef
1km resolution hydrodynamic model and particle tracking simulations. Connectivity data from
24 days in total were used in this investigation to estimate coral larval flows. These data
span six years from 2015 to 2020. The years 2015, 2016, 2018 and 2019 are each represented
by connectivity matrices covering spawning periods over three days in each year (12 days in
total), while 2017 and 2020 are each represented by six spawning days each, coming to a
total of 24 days of connectivity matrices. The values in these matrices indicate the
probability of larval transfer between each pair of reefs in the GBR, herein referred to as
connectivity strengths. Indicative connectivity strengths are implemented in ADRIA-CoralBlox
as the mean of all available data for each location (i.e., mean of all 24 spawning days).
While the connectivity strengths do not vary temporally within each ADRIA-CoralBlox
simulation, they do vary between reefs. This spatial variability enables investigations into
the influence of connectivity strength and its role in driving recovery processes.

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef (excluding retained larvae from the target
reef) and the mean timeseries absolute coral cover (km^2^) of each matching source reef.
This metric provides an approximation of overall larval flow into a reef, with high values
indicating the target reef is supplied high volumes of larvae via strong connections and low
values indicating the target reef has weak (or no) incoming connections or is supplied by
small source populations. A limitation of weighted incoming connectivity is that it does not
incorporate the proportion of the source reef's population that is reproducing and producing
larvae, instead taking the total coral cover area as an indicator of larval productivity.
Outgoing connection strength is calculated as the sum of all outgoing transition
probabilities from a target reef. Outgoing connection strength has a theoretical maximum of
one (all larvae reach settlement on sink reefs), however the large size of the GBR combined
with larval mortality considerations result in outgoing connection strength values much
lower than one. The ratio of source-to-sink connectivity ($O:I$) for a target reef
represents the strength of the reef's outgoing larval flow relative to it's incoming larval
flow. This is calculated as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where S represents the sum of
connectivity probabilities, N represents the number of connections, O denotes outgoing
connectivity from the target reef and I denotes incoming connectivity to the target reef. If
a reef has no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming
connections in this calculation, do not directly represent the larval supply gained as the
breeding populations from which connections are coming from are not represented. A
source-to-sink ratio greater than 1 (a value greater than 0 on a log scale) indicates a reef
has greater outgoing than incoming larval flow. Reef larval retention probability indicates
the proportion of spawned larvae from a target reef that remain to settlement.

@@

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median time series for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour. Assessment metrics used in analyses can be
found in @stbl-metrics.

@@

Time series clustering was applied over the time frame between 2025 and 2075. The clustering
was conducted for three spatial scales: (1) the whole of GBR; (2) each GBR management area;
and (3) for each reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). The analyses were
conducted at the different spatial scales as, conceptually, the temporal behaviour
constituting "low", "medium" or "high" coral cover may differ at the different spatial
scales. As it was, analyses at the whole-of-GBR scale were found to lead to identical
conclusions regarding reef characteristics which influence coral cover and so for brevity
are not shown here.

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, outgoing connectivity, and log10-transformed
outgoing-to-incoming connectivity ratio. The full dataset consisted of the time series
cluster assignment for each of the 2197 reefs under the five GCMs explored, for a total of
10,985 samples. The dataset was randomly split into training (60%) and testing (40%) subsets
to train and assess classification performance. The RF classifier achieved 80% accuracy on
the training set and 60% on the test set.

A 3-fold cross-validation was also conducted, arriving at a similar 59% accuracy rate. F1
performance scores for Low, Medium, and High clusters were determined to be 0.70, 0.43 and
0.64 respectively, with precision of 0.62, 0.51 and 0.66 when evaluated with the test set
(consisting of 4394 samples) (@stbl-confusion-matrix; @stbl-rf-cluster-performance). Lower
performance with the Medium cluster reflects the inherent difficulty of classifying
transitional states in natural systems. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier, while not
accurate, was deemed sufficiently informative for the purpose of assessing the influence of
reef characteristics.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically varied (one-at-a-time) across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. Use of the median predicted
probability was also assessed but was found not to substantially influence the
interpretation of results (and so are not shown here). The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and connectivity metrics, two-way partial dependence analysis
was performed on grids along the ranges of depth and weighted incoming connectivity, and
depth and outgoing connection strength.

@@

In contrast to the broad temporal behaviour at the Management area scale, the relationship
between reef characteristics and assigned clusters were found to be similar at the bioregion
scale across GCMs. Across bioregions, reefs in the High cluster not only had greater levels
of coral recovery but displayed persistence to 2050
(@sfig-access-esm1-5-bioregion-timeseries). Qualitative representations of bioregion scale
clustering results across assessed reef characteristics can be found in the supplementary
material, with Partial Dependence Analysis results displayed in the main text.

@@

Partial Dependence Analysis revealed reef median depth, carrying capacity and mean time
series DHW level have the highest feature importance scores, accounting for 29%, 15% and 15%
predictive importance respectively (@fig-ts-cluster-pdp; Panel A). Area weighted incoming
connectivity, outgoing connectivity strength, source-to-sink connectivity ratio and larval
retention probability account for \~12%, 11%, 10% and 8% respectively.

@@

![Reef Degree Heating Week time series clustered within each management area, for each GCM
analysed. Time series were clustered based on their behaviour between 2030 and 2060. Time
series with comparatively low median area-relative cover is indicated in green, intermediate
cover in orange, and high in blue. Solid lines represent the cluster median time series and
bands represent the 2.5 to 97.5 percentile confidence intervals. Each management area is
represented five times (for each GCM), with each GCM displayed
horizontally.](../figs/GCM_dhw_timeseries_plot.png){#sfig-gcm-dhw-trajectories
fig-align="center" width="14.82cm"}

@@

## Degree Heating Week data processing

@@

Results are only shown and discussed for ACCESS-ESM1-5, an intermediate ECS scenario
(@sfig-ecs). Those interested in results for the other assessed GCMs are referred to Grier
and Iwanaga [-@grier2025]. Reefs in the High cluster were consistently found to have the
greatest median reef depths across bioregions (@sfig-access-esm1-5-bioregion-depth). Across
bioregions, there were no consistent patterns between reef clusters in mean DHW values
(@sfig-access-esm1-5-bioregion-dhwmean), log10-weighted incoming connectivity
(@sfig-access-esm1-5-bioregion-weightedincoming), log10 outgoing connection strength
(@sfig-access-esm1-5-bioregion-outgoingconn), log10 source-to-sink connectivity ratio
(@sfig-access-esm1-5-bioregion-logsosi) and log10 larval retention probability
(@sfig-access-esm1-5-bioregion-selfconn). Across six of the 29 bioregions reefs in the high
cover cluster have lower log10 carrying capacity values than reefs in other clusters
(@sfig-access-esm1-5-bioregion-carrying-capacity).

![Reef area-relative coral cover time series clustered within each bioregion. Time series
were clustered based on their behaviour between 2030 and 2060. Time series with
comparatively low median area-relative cover is indicated in green, intermediate cover in
orange, and high in blue. Solid lines represent the cluster median time series and bands
represent the 95th percentile confidence intervals. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/bioregion_cover_timeseries.png){#sfig-access-esm1-5-bioregion-timeseries
fig-align="center" width="14.82cm"}

![Median depth for reefs within each bioregion under ACCESS-ESM1-5, separated into clusters
assigned by time series clustering. Time series were clustered based on their behaviour
between 2030 and 2060. Time series with comparatively low median area-relative cover is
indicated in green, intermediate cover in orange, and high in blue. Solid lines represent
the median depth for each cluster across reefs and bands represent the distribution of depth
values within each
cluster.](../figs/ACCESS-ESM1-5/bioregion/depth_bioregion_violin.png){#sfig-access-esm1-5-bioregion-depth
fig-align="center" width="14.82cm"}
